Enforced Invariant Confluence

### 2020-01-14 doc/the_vldb_journal_paper/sections/merge_reduction.tex
Fleshed out merge reducibility section.

@@

The proof of \thmref{ReducibilityImpliesIconfluence} is a straightforward
induction. We begin with an \sTIreachable{} expression $e$ and repeatedly
replace any subexpression that merges two merge-free subexpressions with an
equivalent merge-free reachable subexpression (which we can do because $O$ is
merge-reducible). We repeat this process until $e$ has been completely replaced
with an equivalent merge-free reachable expression $e'$. Because $I(s_0)$ and
because our system model only executes transactions that preserve the
invariant, $e'$ (and hence $e$) is guaranteed to satisfy the invariant. Thus,
all reachable states satisfy the invariant, so $O$ is \invariantconfluent{}.

@@

\begin{theorem}\thmlabel{LatticeProperty}
  Given an object $O = (S, \join)$, a start state $s_0 \in S$, a set of
  transactions $T$, and an invariant $I$, if the following criteria are met,
  then $O$ is \sTImergereducible{} (and therefore \sTIconfluent{}).
  \begin{enumerate}
    \item
      $O$ is a join-semilattice.

@@

To show that $O$ is \sTImergereducible, we consider two merge-free reachable
expressions
\[
  e_1 = t_1(\ldots(t_n(s_0))\ldots)
  \quad\text{and}\quad
  e_2 = u_1(\ldots(u_m(s_0))\ldots)
\]
Then, we inductively use criterion (3) with criteria (1) and (2) to show that
\[
  e_3 = t_1(\ldots(
  t_n(u_1(\ldots(u_m(s_0))\ldots))
  )\ldots)
\]
is reachable and evaluates to $e_1 \join e_2$.

@@

% {\input{figures/merge_reducible}}

@@

>>>>

\begin{example}\examplelabel{MergeReducible}
  Consider the distributed object $O = (\ints, \max)$ consisting of integers
  merged by the max function. Our start state state $s_0 = 0$ and our invariant
  $I = \setst{x \in \ints}{x \geq 0}$. Our set $T$ of transactions is the
  infinite set $T = \setst{t_y}{y \in \ints}$ where $t_y(x) = x + y$ is a
  transaction that adds $y$ to the state. For example, $t_2$ is a transaction
  that adds $2$ to the state, and $t_{-3}$ is a transaction which subtracts $3$
  from the state.
  %
  $O$ is \sTImergereducible{}. Consider two merge-free \sTIreachable{}
  expressions $e_1$ and $e_2$ that evaluate to states $x_1$ and $x_2$. Without
  loss of generality, assume $x_1 \geq x_2$. Then, we can replace $e_1 \join
  e_2$ (which evaluates to $x_1$) with $e_1$. We can also replace it with
  $t_{x_1}(0)$.
\end{example}

\begin{example}\examplelabel{NotMergeReducible}
  Consider the distributed object $O = (\set{X \subseteq \ints}, \join)$ in
  which each state is a set of integers and where $X_1 \join X_2 = \set{y}$
  where $y = \sum_{x \in X_1} x + \sum_{x \in X_2} x$. Our start state state
  $s_0 = \emptyset{}$ and our invariant $I = \setst{X}{\forall x \in X.\ x \
  \text{is even}}$. Our set $T$ of transactions is the set $T = \set{t_0, t_2,
  t_4}$ where $t_i(X) = X \cup \set{i}$ is a transaction that adds $i$ to the
  state. For example, $t_2(\set{0}) = \set{0, 2}$.
  %
  $O$ is \emph{not} \sTImergereducible{}. Consider the two merge-free
  \sTIreachable{} expressions $e_1 = t_2(\emptyset)$ and $e_2 =
  t_4(\emptyset)$. $e_1 \join e_2$ evaluates to $\set{6}$, but there does not
  exist a merge-free expression that evaluates to $\set{6}$.
\end{example}

{\input{figures/merging_diagram.tex}}

@@

\begin{proof}
  Intuitively, the proof of \thmref{ReducibilityImpliesIconfluence} is a
  straightforward induction. We begin with an \sTIreachable{} expression $e$
  and repeatedly replace any subexpression that merges two merge-free
  subexpressions with an equivalent merge-free reachable subexpression (which
  we can do because $O$ is merge-reducible). We repeat this process until $e$
  has been completely replaced with an equivalent merge-free reachable
  expression $e'$. Because $I(s_0)$ and because our system model only executes
  transactions that preserve the invariant, $e'$ (and hence $e$) is guaranteed
  to satisfy the invariant. Thus, all reachable states satisfy the invariant,
  so $O$ is \invariantconfluent{}. An illustration of this idea is given in
  \figref{MergingDiagram}.

  More formally, we prove by structural induction on $e$, that for all
  \sTIreachable{} expressions $e$, there exists a merge-free \sTIreachable{}
  expression $e'$ such that $eval(e) = eval(e')$.
  \begin{itemize}
    \item \textbf{Case 1: $e = s_0$.}
      Trivially, $e' = s_0$.

    \item \textbf{Case 2: $e = t(e_1)$.}
      $e_1$ is \sTIreachable{}, so by the inductive hypothesis, there exists a
      merge-free \sTIreachable{} expression $e_1'$ such that $eval(e_1) =
      eval(e_1')$. $t(e_1)$ is \sTIreachable{}, so $I(t(e_1))$. Because
      $eval(e_1) = eval(e_1')$, we know also that $I(t(e_1'))$. Thus, $t(e_1')$
      is \sTIreachable{} (and join free), so we can let $e' = t(e_1')$.

    \item \textbf{Case 3: $e = e_1 \join e_2$.}
      $e_1$ and $e_2$ are \sTIreachable{}, so by the inductive hypothesis,
      there exists equivalent merge-free \sTIreachable{} expressions $e_1'$ and
      $e_2'$. $O$ is \sTImergereducible{}, so there exists an equivalent
      merge-free \sTIreachable{} expression $e'$.
  \end{itemize}

  Consider an arbitrary \sTIreachable{} expression $e$ and it's equivalent
  merge-free \sTIreachable{} counterpart $e'$. $e'$ is either $s_0$ or
  $t(e'')$.  In either case, it satisfies the invariant, so $O$ is
  \sTIconfluent{}.
\end{proof}

Note that while merge-reducibility is a sufficient condition for
\invariantconfluence{}, it is not necessary. The object in
\exampleref{NotMergeReducible} is \invariantconfluent{} but not
merge-reducible.

@@

\begin{theorem}\thmlabel{LatticeProperty}
  Given an object $O = (S, \join)$, a start state $s_0 \in S$, a set of
  transactions $T$, and an invariant $I$, if the following criteria are met,
  then $O$ is \sTImergereducible{} (and therefore \sTIconfluent{}).
  \begin{enumerate}
    \item
      $O$ is a join-semilattice. That is, $\join$ is associative ($(x \join y)
      \join z = x \join (y \join z)$), commutative ($x \join y = y \join x$),
      and idempotent ($x \join x = x$).

@@

\newcommand{\bart}[1]{\overline{t_{#1}}}
\newcommand{\baru}[1]{\overline{u_{#1}}}
\newcommand{\bartu}[2]{\bart{#1}(\baru{#2}(s_0))}
\begin{proof}
  Let
  \[
    \hfill
    e_1 = t_n(t_{n-1}(\ldots (t_1(s_0))\ldots))
    \hfill
  \]
  and
  \[
    \hfill
    e_2 = u_m(u_{m-1}(\ldots (u_1(s_0))\ldots))
    \hfill
  \]
  be two arbitrary merge-free \sTIreachable{} expressions.
  For ease of notation, let
  \[
    \hfill
    \bart{i} = t_i(\ldots(t_1(s_0))\ldots)
    \quad\text{and}\quad
    \baru{j} = u_j(\ldots(u_1(s_0))\ldots)
    \hfill
  \]
  We want to show that there exists some merge-free \sTIreachable{} expression
  that is equivalent to $e_1 \join e_2$.
  %
  To do so, we prove by strong induction on $k \in \nats$ that if $k = i + j$
  where $0 \leq i \leq n$ and $0 \leq j \leq m$, $\bartu{i}{j}$ is
  \sTIreachable{} and $eval(\bartu{i}{j}) = eval(\bart{i}(s_0) \join
  \baru{j}(s_0))$.
  \begin{itemize}
    \item \textbf{Case $k = 0$.}
      $i = j = 0$, so $\bartu{0}{0} = s_0$ which is trivially \sTIreachable{}
      and equivalent to $\bart{0}(s_0) \join \baru{0}(s_0) = s_0 \join s_0$
      which evaluates to $s_0$ (because $\join$ is idempotent).

    \item \textbf{Case $k = 1$.}
      Without loss of generality, assume $i = 1$ and $j = 0$. Then,
      $
        \bartu{1}{0} = t_1(s_0)
      $ which is \sTIreachable{} because it is a subexpression of $\bart{n}$
      which is \sTIreachable{}. Moreover, it is equivalent to $\bart{1}(s_0)
      \join \baru{0}(s_0) = t_1(s_0) \join s_0$ which evaluates to $s_{t_1}
      \join s_0 \join s_0 = s_{t_1} \join s_0 = t_1(s_0)$ for some $s_{t_1} \in
      S$.

    \item \textbf{Case $k \geq 2$.}
      If $i = 0$, then $j = k$ and $\baru{k}(s_0)$ is \sTIreachable{} because
      it is a subexpression of $\baru{m}(s_0)$. Also, it is equivalent to
      $\bart{0}(s_0) \join \baru{k}(s_0)$ which evaluates to $\baru{k}(s_0)$.
      %
      The symmetric result holds if $j = 0$.

      Otherwise, $i, j > 1$. Let
      \begin{align*}
        e_{i-1,j-1} &= \bartu{i-1}{j-1} \\
        e_{i,j-1} &= \bartu{i}{j-1} \\
        e_{i-1,j} &= \bartu{i-1}{j}
      \end{align*}
      By the inductive hypothesis, $e_{i-1,j-1}$, $e_{i,j-1}$, and $e_{i-1,j}$
      are all \sTIreachable{}. By condition 3 (with $s = eval(e_{i-1,j-1})$,
      $t_1$ = $t_i$, and $t_2 = u_j$), $I(e_{i,j-1} \join e_{i-1,j})$.
      $e_{i,j-1} \join e_{i-1,j} = t_i(e_{i-1,j}) = u_j(e_{i,j-1}) =
      \bartu{i}{j}$, so $I(\bartu{i}{j})$. Therefore, $\bartu{i}{j}$ is
      \sTIreachable{}.
  \end{itemize}
\end{proof}

{\input{figures/diamond_proof_diagram.tex}}

An illustration of this proof is given in \figref{DiamondProofDiagram}. We
arrange the expressions $e_1$ and $e_2$ as the left and top edges of a square
grid. Each point in the grid represents a state (with $s_0$ in the top left
corner), and each edge represents the application of a transaction. A state is
circled if we know it satisfies the invariant.
%
Condition (1) and (2) tell us that the order in which we apply transactions is
immaterial. Thus, if we begin at the top left of the square and walk to any
other point in the square, applying transactions along the way, it does not
matter which path we take. They are all equivalent. Condition (4) tells us that
the top-left corner satisfies the invariant. We induct to repeatedly apply
condition (3) to ``fill in'' the square, one block at a time. In iteration $k$,
we discover that all points with a Manhattan distance of $k$ from the top left
corner satisfy the invariant. Ultimately, we conclude that the bottom left
corner (i.e., $e_1 \join e_2$) satisfies the invariant and is equivalent to
$\bartu{n}{m}$.

@@

{\input{figures/merge_reducible}}

@@

The relationship between \invariantconfluence{}, \invariantclosure{},
merge-reducibility, and criteria (1)-(4) is illustrated in
\figref{IclosureVsReducible}.
